---
title: "Querying MIMIC-III -- Tutorial Problem -- Tidyverse/dplyr"
output:
  html_document:
    toc: yes
  html_notebook:
    toc: yes
---

efg | 2018-09-21 | Updated 2018-12-30.

# Overview

The purpose of this script is to reproduce the "Tutorial Problem" from [Querying MIMIC-III](https://mimic.physionet.org/tutorials/intro-to-mimic-iii/), Section 8, using Tidyverse/dplyr in the examples instead of SQL used in the original solution.  This enables a direct comparison between solving the problem using SQL and dplyr/dbplyr.

The examples from Section 3 “Patient numbers” through Section 7 “Services are in a separate notebook.

Many of the comments below are from the source above.

Prerequisites: This tutorial assumes that you have an active connection to an instance of MIMIC-III running on PostgreSQL. Note also that all the queries are written assuming that the MIMIC-III database is on your default search path. 

NOTES:

* RPostgreSQL works fine with dates before 1970, but seems to truncate datetime timestamps to dates in RStudio, and does not compute time intervals correctly at times.

* Rpostgres does NOT work with dates before 1970 (*fix expected in Feb. 2019*), but is needed for Step 7 here for results to match between SQL and dplyr solutions.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

Packages

```{r, comment=NA}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(dbplyr,    warn.conflicts=FALSE)
library(lubridate, warn.conflicts=FALSE)   # date computations
library(kableExtra)
```

Helper function:  Common formatting mostly for data.frames/tibbles below

```{r}
Show <- function(data, caption = NULL)
{
  data                               %>%
  kable("html", caption = caption)   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE, font_size=12)
}
```        

Open database

```{r, comment=NA}
MimicDB <- dbConnect(RPostgreSQL::PostgreSQL(),    # PostgreSQL for Steps 1 - 6
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     options  = "-c search_path=mimiciii")

MimicVersion <- "MIMIC-III, v1.4"
```

Table references

```{r, comment=NA}
icustays   <- tbl(MimicDB, in_schema("mimiciii", "icustays"))
patients   <- tbl(MimicDB, in_schema("mimiciii", "patients"))
admissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
```

# Querying MIMIC-III

## 8. Tutorial problem

How would you gather useful information about patients admitted to the ICU? The problem can be broken down into several parts:

### Step 1

First start with retrieving `subject_id`, `hadm_id`, `icustay_id`, `intime`, and `outtime` from the `icustays` table.

```{r, comment=NA}
results1 <-
  icustays                                                  %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime)  %>%
  collect()    # force execution by DB server and return as R object
```

All results returned to in-memory tibble

```{r, comment=NA}
dim(results1)
```

```{r, comment=NA}
results1 %>% 
  head() %>% 
  Show("Selected fields from icustays")
```

### Step 2

Using the patients table retrieve the calculated age of patients.

```{r, comment=NA}
results2 <-
  icustays                                                             %>%
  inner_join(patients, by="subject_id")                                %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, dob)        %>%
  collect()                                                            %>%
  mutate(ageYears = (as.numeric(floor_date(intime, unit="day") - 
                                floor_date(dob,    unit="day"), 
                                units="days")  /  365.242)  %>%  round(2))  %>%
  select(-dob)
```

```{r, comment=NA}
dim(results2)
```

```{r, comment=NA}
results2   %>% 
  head()   %>%
  Show(caption = "Calculated ages")
```

The following graphic is an addition to the original example.

```{r, comment=NA, fig.width=8, fig.height=6}
ggplot(results2, aes(x=ageYears, y=..density..))            +
  geom_histogram(fill="cornsilk", color="grey80", bins=50)  +
  geom_density()                                            +
  labs(title    = "Age density",
       subtitle = paste0(nrow(results2), " records"),
       x = "Age",
       y = "Density",
       caption = MimicVersion)
         
```

Note the strange convention for ages > 89 to be represented as ~300 years.

### Step 3

Separate neonates from adult patients.

```{r, comment=NA}
results3 <-
  icustays                                                            %>%
  inner_join(patients, by="subject_id")                               %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, dob)       %>%
  collect()                                                           %>%
  mutate(ageYears = (as.numeric(floor_date(intime, unit="day") - 
                                floor_date(dob,    unit="day"), 
                                units="days")  /  365.242)  %>%  round(2)) %>%
  select(-dob)                                                        %>%
  mutate(icuStayAgeGroup =
           case_when(
             ageYears <=  1 ~ "neonate",
             ageYears <= 14 ~ "middle",
             ageYears > 100 ~ ">89",
             TRUE           ~ "adult"
           ))                       
```

```{r, comment=NA}
dim(results3)
```

```{r, comment=NA}
results3 %>% head() %>% Show()
```

```{r, comment=NA}
results3                    %>%
  group_by(icuStayAgeGroup) %>%
  count()                   %>%
  Show(caption = "Counts by Age Group")
```

### Step 4

By incorporating the admissions table, find how long each stay was BEFORE the patients were admitted to the ICU.

Changed original query to return preiculos (pre-ICU Length-of-Stay) in days instead of years.

```{r, comment=NA}
results4 <-
  icustays                                                            %>%
  inner_join(patients,   by = "subject_id")                           %>%
  inner_join(admissions, by = c("subject_id", "hadm_id"))             %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, 
         dob, admittime)                                              %>%
  collect()                                                           %>%
  mutate(ageYears      = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(dob,       unit="day"), 
                                     units="days")  / 365.242)  %>%  round(2),
         
         preiculosDays = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(admittime, unit="day"), 
                                     units="days")) 
         
        )                                                             %>%
  select(-dob, -admittime)                                            %>%
  mutate(icuStayAgeGroup =
           case_when
           (
             ageYears <=  1 ~ "neonate",
             ageYears <= 14 ~ "middle",
             ageYears > 100 ~ ">89",
             TRUE           ~ "adult"
           )
         )                       
```

```{r, comment=NA}
dim(results4)
```

```{r, comment=NA}
results4 %>% head() %>% Show(caption = "Pre-ICU Length of Stay [days]")
```
```{r, comment=NA}
counts    <- table(results4$icuStayAgeGroup, useNA = "ifany")
facetName <- paste0(names(counts), " [", counts, "]")

variableLabeller <- function(variable,value)
{
  return(facetName[value])
}
```

```{r, fig.height=8, fig.width=8}
ggplot(results4, aes(x = preiculosDays, y= ..density..))        +
  geom_histogram(fill = "cornsilk", color = "grey80", bins = 50)   +
  geom_density()                                                   +  
  xlim(0, 7)                                                       +
  facet_wrap(~icuStayAgeGroup, nrow=length(counts), 
             labeller = as_labeller(variableLabeller))



```

### Step 5

Next find the date of the patient’s death if applicable.

```{r, comment=NA}
results5 <-
  icustays                                                            %>%
  inner_join(patients,   by = "subject_id")                           %>%
  inner_join(admissions, by = c("subject_id", "hadm_id"))             %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, 
         deathtime, dob, admittime)                                   %>%
  collect()                                                           %>%
  mutate(ageYears      = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(dob,       unit="day"), 
                                     units="days")  / 365.242)  %>%  round(2),
         
         preiculosDays = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(admittime, unit="day"), 
                                     units="days")) 
         
        )                                                             %>%
  select(-dob, -admittime)                                            %>%
  mutate(icuStayAgeGroup =
           case_when
           (
             ageYears <=  1 ~ "neonate",
             ageYears <= 14 ~ "middle",
             ageYears > 100 ~ ">89",
             TRUE           ~ "adult"
           )
         )                       
```

```{r, comment=NA}
dim(results5)
```

```{r, comment=NA}
results5 %>% head() %>% Show(caption = "Show Death Time (if applicable)")
```

```{r}
deaths <-
  results5                 %>%
  filter(!is.na(deathtime)) 
```

Counts of deaths

```{r}
deaths %>% count() %>% Show()
```

```{r}
deaths                                              %>% 
  group_by(icuStayAgeGroup)                         %>%
  summarize(n = n(), 
            minAge    = min(ageYears),
            Q25age    = quantile(ageYears, 0.25),
            medianAge = median(ageYears),
            meanAge   = round(mean(ageYears), 2),
            Q75age    = quantile(ageYears, 0.75),
            maxAge    = max(ageYears))             %>%
  Show(caption = "Stats by age group for patients who died")
```

### Step 6

Then find those deaths that occurred while the patients were in the hospital.

```{r, comment=NA}
results6 <-
  icustays                                                                %>%
  inner_join(patients,   by = "subject_id")                               %>%
  inner_join(admissions, by = c("subject_id", "hadm_id"))                 %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, 
         deathtime, dob, admittime, hospital_expire_flag)                 %>%
  collect()                                                               %>%
  mutate(ageYears      = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(dob,       unit="day"), 
                                     units="days")  / 365.242)  %>%  round(2),
         
         preiculosDays = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(admittime, unit="day"), 
                                     units="days")),
         icuStayAgeGroup =
           case_when
           (
             ageYears <=  1 ~ "neonate",
             ageYears <= 14 ~ "middle",
             ageYears > 100 ~ ">89",
             TRUE           ~ "adult"
           ),
         
         hospitalExpireFlag = ifelse(hospital_expire_flag == 1, "Y", "N")
        )                                                                %>%
  
  select(-dob, -admittime, -hospital_expire_flag)
```

```{r, comment=NA}
dim(results6)
```

```{r, comment=NA}
results6 %>% head() %>% Show()
```

```{r}
results6                                             %>%
  group_by(icuStayAgeGroup, hospitalExpireFlag)      %>%
  count()                                            %>%
  spread(hospitalExpireFlag, n)                      %>%
  Show(caption = "Counts of Hospital Deaths ('Y')")
```

### Step 7 

```{r, comment=NA}
dbDisconnect(MimicDB)
```

```{r}
MimicDB <- dbConnect(RPostgres::Postgres(),    # Postgres for Step 7
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     options  = "-c search_path=mimiciii")
```

```{r, comment=NA}
icustays   <- tbl(MimicDB, in_schema("mimiciii", "icustays"))
patients   <- tbl(MimicDB, in_schema("mimiciii", "patients"))
admissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
```

```{r, comment=NA}
results7 <-
  icustays                                                                %>%
  inner_join(patients,   by = "subject_id")                               %>%
  inner_join(admissions, by = c("subject_id", "hadm_id"))                 %>%
  select(subject_id, hadm_id, icustay_id, intime, outtime, 
         deathtime, dob, admittime, hospital_expire_flag, 
         dischtime, discharge_location)                                   %>%
  collect()                                                               %>%
  mutate(ageYears      = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(dob,       unit="day"), 
                                     units="days")  / 365.242)  %>%  round(2),
         
         preiculosDays = (as.numeric(floor_date(intime,    unit="day") - 
                                     floor_date(admittime, unit="day"), 
                                     units="days")),
         icuStayAgeGroup =
           case_when
           (
             ageYears <=  1 ~ "neonate",
             ageYears <= 14 ~ "middle",
             ageYears > 100 ~ ">89",
             TRUE           ~ "adult"
           ),
         
         hospitalExpireFlag = ifelse(hospital_expire_flag == 1, "Y", "N"),
         
         icuStayExpireFlag = 
           case_when
           (
             (deathtime >= intime) &   # [don't use between](https://wiki.postgresql.org/wiki/Don%27t_Do_This)
             (deathtime <= outtime)                   ~ "Y",  
             
             deathtime <= intime                      ~ "Y",
             
             (dischtime <= outtime) & 
             (discharge_location == "DEAD/EXPIRED")   ~ "Y",
               
             TRUE                                     ~ "N"
           )
        )                                                                %>%
  
  select(-dob, -admittime, -hospital_expire_flag, -dischtime, -discharge_location) %>%
  arrange(subject_id, hadm_id)
```

```{r, comment=NA}
dim(results7)
```

```{r, comment=NA}
results7 %>% head() %>% Show()
```

```{r}
table(results7$icuStayExpireFlag)
```

```{r}
results7                                          %>%
  group_by(icuStayAgeGroup, icuStayExpireFlag)    %>%
  count()                                         %>%
  spread(icuStayExpireFlag, n)                    %>%
  Show(caption = "Counts of Deaths in ICU ('Y')")
```

```{r, comment=NA}
table(results7$hospitalExpireFlag, results7$icuStayExpireFlag)
```

# Close database

```{r, comment=NA}
dbDisconnect(MimicDB)
```

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
