---
title: "Querying MIMIC-III -- Tutorial Problem -- SQL"
output:
  html_document:
    toc: yes
  html_notebook:
    toc: yes
---

efg | 2018-09-21 | Updated 2018-12-16.

# Overview

The purpose of this RStudio notebook is to reproduce the SQL "Tutorial Problem" from [Querying MIMIC-III](https://mimic.physionet.org/tutorials/intro-to-mimic-iii/), Section 8.  Unfortuantely, the online source shows SQL code but does not show the “correct” answers, so a definitive conclusion that this all matches is not possible.

The examples from Section 3 “Patient numbers” through Section 7 “Services are in a separate notebook.

**An equivalent Tidyverse example shows roughly the same steps using RStudio notebooks and dplyr/dbplyr for data manipulation instead of SQL.**  This enables a direct comparison between solving the problem using SQL and dplyr/dbplyr.

Many of the comments below are from the source above.

SQL statements were lightly edited for readability in the SQL program "chunks".  The output of each SQL statement was directed to the variable `output.var` in the SQL chunk definition to document the dimensions of the returned results.  `output.var` results could be displayed or manipulated in any way.

Prerequisites: This tutorial assumes that you have an active connection to an instance of MIMIC-III running on PostgreSQL. Note also that all the queries are written assuming that the MIMIC-III database is on your default search path. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

# Setup

```{r, comment=NA}
library(DBI)
library(RPostgres)
```

```{r}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     host     = "localhost",
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",   # See RBloggers 2018-03-16 posting
                     options  = "-c search_path=mimiciii")
```

# Querying MIMIC-III

## 8. Tutorial problem

How would you gather useful information about patients admitted to the ICU? The problem can be broken down into several parts:

### Step 1

First start with retrieving `subject_id`, `hadm_id`, `icustay_id`, `intime`, and `outtime` from the `icustays` table.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime
FROM   
  icustays ie
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 2

Using the patients table retrieve the calculated age of patients.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime,
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) AS age
FROM   
  icustays ie

  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id;
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 3

Separate neonates from adult patients.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime,
  
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) AS age,
  
  CASE
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 1
    THEN 'neonate'
    
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 14
    THEN 'middle'
    
    -- all ages > 89 in the database were replaced with 300
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) > 100
    then '>89'
    
    ELSE 'adult'
  END AS ICUSTAY_AGE_GROUP
  
FROM 
  icustays ie

  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 4

By incorporating the admissions table, find how long each stay was BEFORE the patients were admitted to the ICU.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime,
  
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))      /365.242, 2) as age,
  ROUND((cast(ie.intime as date) - cast(adm.admittime as date))/365.242, 2) as preiculos,
  
  CASE
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 1
    THEN 'neonate'
    
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 14
    THEN 'middle'
    
    -- all ages > 89 in the database were replaced with 300
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) > 100
    THEN '>89'
    
    ELSE 'adult'
  END AS ICUSTAY_AGE_GROUP
        
FROM 
  icustays ie
  
  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id
  
  INNER JOIN admissions adm
  ON ie.hadm_id = adm.hadm_id
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 5

Next find the date of the patient’s death if applicable.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime, 
  adm.deathtime,
  
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))      /365.242, 2) as age,
  ROUND((cast(ie.intime as date) - cast(adm.admittime as date))/365.242, 2) AS preiculos,
  
  CASE
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 1
    THEN 'neonate'
    
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 14
    THEN 'middle'
    
    -- all ages > 89 in the database were replaced with 300
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) > 100
    THEN '>89'
    
    ELSE 'adult'
  END AS ICUSTAY_AGE_GROUP
      
FROM 
  icustays ie
  
  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id
  
  INNER JOIN admissions adm
  ON ie.hadm_id = adm.hadm_id
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 6

Then find those deaths that occurred while the patients were in the hospital.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime, 
  adm.deathtime,
    
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))      /365.242, 2) AS age,
  ROUND((cast(ie.intime as date) - cast(adm.admittime as date))/365.242, 2) AS preiculos,
  
  CASE
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 1
    THEN 'neonate'
      
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 14
    THEN 'middle'
      
    -- all ages > 89 in the database were replaced with 300
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) > 100
    THEN '>89'
      
    ELSE 'adult'
  END AS ICUSTAY_AGE_GROUP,
      
  -- note that there is already a "hospital_expire_flag" field in the admissions table which you could use
  CASE
    WHEN adm.hospital_expire_flag = 1 
    THEN 'Y'
    
    ELSE 'N'
  END AS hospital_expire_flag
    
FROM 
  icustays ie
  
  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id
  
  INNER JOIN admissions adm
  ON ie.hadm_id = adm.hadm_id
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

### Step 7

Find how many of those deaths occurred within the ICU.

```{sql, connection=MimicDB, output.var="SQLresults"}
SELECT 
  ie.subject_id, 
  ie.hadm_id, 
  ie.icustay_id,
  ie.intime, 
  ie.outtime, 
  adm.deathtime,
    
  ROUND((cast(ie.intime as date) - cast(pat.dob as date))      /365.242, 2) AS age,
  ROUND((cast(ie.intime as date) - cast(adm.admittime as date))/365.242, 2) AS preiculos,
  
  CASE
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 1
    THEN 'neonate'
    
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) <= 14
    THEN 'middle'
    
    -- all ages > 89 in the database were replaced with 300
    WHEN ROUND((cast(ie.intime as date) - cast(pat.dob as date))/365.242, 2) > 100
    THEN '>89'
    
    ELSE 'adult'
  END AS ICUSTAY_AGE_GROUP,

  -- note that there is already a "hospital_expire_flag" field in the admissions table which you could use
  CASE
    WHEN adm.hospital_expire_flag = 1 
    THEN 'Y'           
    ELSE 'N'
  END AS hospital_expire_flag,

  -- note also that hospital_expire_flag is equivalent to "Is adm.deathtime not null?"
  CASE
    WHEN adm.deathtime BETWEEN ie.intime and ie.outtime
    THEN 'Y'
    
    -- sometimes there are typographical errors in the death date, so check before intime
    WHEN adm.deathtime <= ie.intime
    THEN 'Y'
    
    WHEN adm.dischtime          <= ie.outtime     AND
         adm.discharge_location  = 'DEAD/EXPIRED'
    THEN 'Y'
    
    ELSE 'N'
  END AS ICUSTAY_EXPIRE_FLAG

FROM 
  icustays ie

  INNER JOIN patients pat
  ON ie.subject_id = pat.subject_id
  
  INNER JOIN admissions adm
  ON ie.hadm_id = adm.hadm_id
```

```{r, comment=NA}
dim(SQLresults)
head(SQLresults)
```

# Close database

```{r, comment=NA}
dbDisconnect(MimicDB)
```

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
