---
title: "Download MIMIC-III Files"
output: html_notebook
---

efg | 2018-07-03

Start at [PhysioNet web page](https://physionet.org/) and select  [MIMIC-III Database](https://mimic.physionet.org/).

Select **Data** and complete all the requirements to access MIMIC-III.

Once you have a username and password, run this script ...

Assume *scripts* starting directory.

```
    MIMIC-III
        scripts
            000-Download
                DownloadMIMIC-III.Rmd
        raw
        data
```

A *raw* directory will be created for the downloadeded compressed files, which will be unzipped to form the *data* directory.

THe *raw* directory could be deleted at the conclusion of this script, but that is left as a manual operation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()
```

```{r}
suppressPackageStartupMessages( library(R.utils) )   # downloadFile, gunzip
suppressPackageStartupMessages( library(dplyr) )
library(tools)      # md5sum
library(tibble)     # rownames_to_column
library(kableExtra) # kable_styling
```

```{r}
rawDIR  <- "../../raw/"
if (!dir.exists(rawDIR)) dir.create(rawDIR)

dataDIR <- "../../data/" 
if (!dir.exists(dataDIR)) dir.create(dataDIR)
```

## Prompts for the PhysioNet username and password

```{r}
username <- rstudioapi::showPrompt(title="PhysioNet Username:", message="Username", default="")
password <- rstudioapi::askForPassword(prompt="PhysioNet Password:")  # hide password 
```

## Download files

Somewhat tricky because of password-protected site.

I normally use **download** from **downloader** package, but it offers no password option.  **downloadFile** does the trick.

### md5sum files

```{r}
URL <- "https://physionet.org/works/MIMICIIIClinicalDatabase/files/version_1_4/"

filename <- "checksum_md5_zipped.txt"
downloadFile(paste0(URL, filename), mode="wb", username=username, password=password, quiet=TRUE)

filename <- "checksum_md5_unzipped.txt"
downloadFile(paste0(URL, filename), mode="wb", username=username, password=password, quiet=TRUE)
```

### gz files

List of files to download to "raw" directory.

```{r}
gzFileNames <- c(
  "ADMISSIONS.csv.gz",
  "CALLOUT.csv.gz",
  "CAREGIVERS.csv.gz",
  "CHARTEVENTS.csv.gz",
  "CPTEVENTS.csv.gz",
  "DATETIMEEVENTS.csv.gz",
  "DIAGNOSES_ICD.csv.gz",
  "DRGCODES.csv.gz",
  "D_CPT.csv.gz",
  "D_ICD_DIAGNOSES.csv.gz",
  "D_ICD_PROCEDURES.csv.gz",
  "D_ITEMS.csv.gz",
  "D_LABITEMS.csv.gz",
  "ICUSTAYS.csv.gz",
  "INPUTEVENTS_CV.csv.gz",
  "INPUTEVENTS_MV.csv.gz",
  "LABEVENTS.csv.gz",
  "MICROBIOLOGYEVENTS.csv.gz",
  "NOTEEVENTS.csv.gz",
  "OUTPUTEVENTS.csv.gz",
  "PATIENTS.csv.gz",
  "PRESCRIPTIONS.csv.gz",
  "PROCEDUREEVENTS_MV.csv.gz",
  "PROCEDURES_ICD.csv.gz",
  "SERVICES.csv.gz",
  "TRANSFERS.csv.gz")
```

```{r}
for (i in 1:length(gzFileNames))
{
  filename <- gzFileNames[i]
  downloadFile(paste0(URL, filename), path=rawDIR,
               mode="wb", username=username, password=password, quiet=TRUE)
}
```

## Get reference .gz md5sums

```{r}
refgzmd5sums <-
  readLines("checksum_md5_zipped.txt") %>%
  tibble()            %>%
  mutate(filename=substr(., 35, nchar(.)), 
         Refmd5sum=substr(., 1, 32))   %>%
         select(-.)
```

## verify md5sums of .gz files

```{r}
gzmd5sums <-
  data.frame( md5sum=md5sum( dir(rawDIR, full.names=TRUE) ) ) %>%
  tibble::rownames_to_column(var="filename")                  %>%
  mutate(filename=substr(filename,11,nchar(filename)))
```

match is OK if md5sum is same as the reference file

```{r}
gzmd5sumMatch <-
  left_join(refgzmd5sums, gzmd5sums, by="filename") %>%
  mutate(match=ifelse(Refmd5sum == md5sum, "OK", "MISMATCH"))

gzmd5sumMatch    %>%
  kable("html")  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                full_width=FALSE, position="left")
```

## ungzip compressed files into data directory

```{r}
for (filename in dir(rawDIR))
{
  cat(filename, "\n")
  gunzip(paste0(rawDIR, filename),
         destname=paste0(dataDIR, substr(filename, 1, nchar(filename)-3)),
         remove=FALSE)
}
```

## Get reference .csv md5sums

```{r}
refcsvmd5sums <-
  readLines("checksum_md5_unzipped.txt")   %>%
  tibble()                                 %>%
  mutate(filename=substr(., 35, nchar(.)),
         Refmd5sum=substr(., 1, 32))       %>%
         select(-.)
```

## verify md5sums of .csv files

```{r}
csvmd5sums <-
  data.frame( md5sum=md5sum( dir(dataDIR, full.names=TRUE) ) ) %>%
  tibble::rownames_to_column(var="filename")                   %>%
  mutate(filename=substr(filename,12,nchar(filename)))
```

```{r}
csvmd5sumMatch <-
  left_join(refcsvmd5sums, csvmd5sums, by="filename") %>%
  mutate(match=ifelse(Refmd5sum == md5sum, "OK", "MISMATCH"))

csvmd5sumMatch    %>%
  kable("html")   %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                full_width=FALSE, position="left")
```

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
