---
title: "MIMIC-III SQL Examples by Pallab Paul -- with dplyr equivalents"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style>                          

efg | 2019-03-03

# Purpose

The purpose of this notebook is to 

* Recreate Postgres `SQL` examples from Pallab Paul's Jupyter notebook in an R Studio notebook.  

See blog post [Predictions using the MIMIC-III Database (Part 1)](https://medium.com/intel-student-ambassadors/predictions-using-the-mimic-iii-database-part-1-1aa58475da40) and details in Jupyter notebook, `mimiciii-sql.ipynb`, which can be [found here](https://github.com/PallabPaul/mimic-mortality-predictions).

* Create equivalent `dplyr` examples in R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
options(width=100)
```

# Setup

Simple function using a small part of info from [Create Awesome HTML Table with knitr::kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).

## Helper function

Common formatting mostly for data.frames/tibbles

```{r, comment=NA}
Show <- function(data, caption="", bigMark="", ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)
}
```     

## Packages

```{r, comment=NA}
library(tidyverse)
library(DBI)

library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(lubridate, warn.conflicts=FALSE)   # date computations, e.g., floor_date
```

Which Postgres package?  There have been some problems with both, but either works fine for the examples below.

```{r, comment=NA}
library(RPostgreSQL)
#library(RPostgres)
```

## Open MIMIC-III

```{r, comment=NA}
#MimicDB <- dbConnect(RPostgreSQL::PostgreSQL(),   # Only shows dates with admittime
MimicDB <- dbConnect(RPostgres::Postgres(),        # Shows dates and times with admittime
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     options  = "-c search_path=mimiciii",
                     bigint   = "integer")   # 64-bit ints not needed (RPostgres Only)
```

## Table references for dplyr

```{r, comment=NA}
admissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
patients   <- tbl(MimicDB, in_schema("mimiciii", "patients"))
transfers  <- tbl(MimicDB, in_schema("mimiciii", "transfers"))
services   <- tbl(MimicDB, in_schema("mimiciii", "services"))
```

# Example 1

## SQL

```{sql, connection=MimicDB, output.var="SQLresults"}   
select count(subject_id)
from   mimiciii.patients
```

```{r, comment=NA}
SQLresults  %>% Show(bigMark = ",")
```

## dplyr

```{r, comment=NA}
patients                  %>%
  summarize(count = n())  %>%
  Show(bigMark = ",")
```

Look at counts after join

```{r, comment=NA}
admissions                                         %>%
  inner_join(patients, by = "subject_id")          %>%
  summarize(nAdmissions = n(),
            nSubjects   = n_distinct(subject_id))  %>%
  Show(bigMark = ",")
```

# Example 2

Paul: *We first start by figuring out the patient’s birth date and the patient’s
admission dates to the ICU.*

## SQL

```{sql, connection=MimicDB, output.var="SQLresults"}   
SELECT p.subject_id, 
       p.dob, 
       a.hadm_id,
       a.admittime, 
       p.expire_flag
FROM   mimiciii.admissions a
INNER JOIN mimiciii.patients p
ON     p.subject_id = a.subject_id
```

```{r, comment=NA}
dim(SQLresults)
```

```{r, comment=NA}
SQLresults %>%
  head()   %>%
  Show()
```

```{r, comment=NA}
SQLresults %>%
  tail()   %>%
  Show()
```

## dplyr

```{r, comment=NA}
dplyrResults <-
  admissions                                                %>%
  inner_join(patients, by = "subject_id")                   %>%
  select(subject_id, dob, hadm_id, admittime, expire_flag)  %>%
  collect()
```

```{r, comment=NA}
dim(dplyrResults)
```

```{r, comment=NA}
dplyrResults %>%
  head()   %>%
  Show()
```

```{r, comment=NA}
dplyrResults %>%
  tail()     %>%
  Show()
```

# Example 3

Paul:  *Next, we figure out the patient's first admission date so that we do not
have multiple records of the same patient and so that we only have one
age per patient*

## SQL

```{sql, connection=MimicDB, output.var="SQLresults"}  
SELECT p.subject_id, 
       p.dob, 
       a.hadm_id,
       a.admittime, 
       p.expire_flag,
       MIN (a.admittime) OVER (PARTITION BY p.subject_id) AS first_admittime
FROM   mimiciii.admissions a
INNER JOIN mimiciii.patients p
ON    p.subject_id = a.subject_id
ORDER BY a.hadm_id, 
         p.subject_id
```

```{r, comment=NA}
dim(SQLresults)
```

```{r, comment=NA}
SQLresults %>%
  head()   %>%
  Show()
```

```{r, comment=NA}
SQLresults %>%
  tail()   %>%
  Show()
```

BUT, there are multiple records for some patients since the result had 58,976 records (# of admissions) instead of 46,520 records (# of patients).

Patients with multiple records had the same `dob` and `first_admission` (age).

[The group by `subject`, `dob` and `gender` eliminates the problem in the next example.]

```{r, comment=NA}
SQLresults                        %>%
  group_by(subject_id)            %>%
  summarize(RecordsPerSubject = n())   %>%
  ungroup()                       %>%
  filter(RecordsPerSubject > 1)   %>%  
  group_by(RecordsPerSubject)     %>%
  count()                         %>%
  ungroup(RecordsPerSubject)      %>%
  Show()
```

```{r, comment=NA}
SQLresults                             %>%
  group_by(subject_id)                 %>%
  summarize(RecordsPerSubject = n())   %>%
  ungroup()                            %>%
  filter(RecordsPerSubject == 15)      %>%  
  Show()
```

```{r, comment=NA}
SQLresults                     %>%
  filter(subject_id == 41976)  %>%
  arrange(admittime)           %>%
  Show()
```

A possible solution to get one record per patient requires only records with `hadm_id` for `first_admittime`:

```{r, comment=NA}
SQLresults                              %>%
  filter(admittime == first_admittime)  %>%
  count()                               %>%
  Show()
```

## dplyr

This dplyr example does what I think the above query was intended to do.

From StackOverFlow: [Select first and last row from grouped data](https://stackoverflow.com/questions/31528981/select-first-and-last-row-from-grouped-data).

```{r, comment=NA}
dplyrResults <-
  admissions                                                %>%
  inner_join(patients, by = "subject_id")                   %>%
  select(subject_id, dob, hadm_id, admittime, expire_flag)  %>%
  
  group_by(subject_id)       %>%
  arrange(admittime)         %>%
  filter(row_number() == 1)  %>%
  ungroup()                  %>%
  rename(first_admittime = admittime)  %>%   

  collect()                   %>%
  arrange(hadm_id, subject_id)
```

There is no need for `admittime` and `first_admittime` since both are the same here.

```{r, comment=NA}
dim(dplyrResults)
```

```{r, comment=NA}
dplyrResults %>%
  head()     %>%
  Show()
```

```{r, comment=NA}
dplyrResults %>%
  tail()     %>%
  Show()
```

There is now only a single record for `subject_id` 41976, instead of 15 records in the SQL version above.

```{r, comment=NA}
dplyrResults                   %>%
  filter(subject_id == 41976)  %>%
  Show()
```

# Example 4

Paul:  *We then find the age of the patient by finding the difference between
their date of birth and the date of their first admission. We put these
age groups into three categories: neonatal (< 15 years of age), adult
(age range of 15–89) and >89 years of age.*

## SQL

```{sql, connection=MimicDB, output.var="SQLresults"}  
WITH first_admission_time AS
(
  SELECT
      p.subject_id, 
      p.dob, 
      p.gender,
      MIN (a.admittime) AS first_admittime,
      MIN( ROUND( (cast(admittime as date) - cast(dob as date)) / 365.242,2) ) AS first_admit_age
  FROM mimiciii.patients p
  INNER JOIN mimiciii.admissions a
  ON p.subject_id = a.subject_id
  GROUP BY p.subject_id, p.dob, p.gender
  ORDER BY p.subject_id
)
SELECT
    subject_id, 
    dob, 
    gender, 
    first_admittime, 
    first_admit_age,
    CASE
        -- all ages > 89 in the database were replaced with 300
        WHEN first_admit_age > 89
             then '>89'
        WHEN first_admit_age >= 14
             THEN 'adult'
        WHEN first_admit_age <= 1
             THEN 'neonate'
        ELSE 'middle'
    END AS age_group
FROM first_admission_time
ORDER BY subject_id
```

```{r, comment=NA}
dim(SQLresults)
```

```{r, comment=NA}
SQLresults  %>%
  head()    %>%
  Show()
```

## dplyr

I'm not sure how to use the Postgres `cast` and `as` through `dbplyr`.

Let's use database to extract data, and complete the computations in R.

```{r, comment=NA}
dplyrResults <-
  admissions                                  %>%
  inner_join(patients, by = "subject_id")     %>%
  select(subject_id, 
         dob, 
         hadm_id, 
         admittime, 
         gender)                              %>%
  
  group_by(subject_id, gender)  %>%
  arrange(admittime)            %>%
  filter(row_number() == 1)     %>%
  ungroup()                     %>%
  
  rename(first_admittime = admittime)  %>%   
  arrange(subject_id)           %>%
  collect()                     %>%  # extract from database
  
  mutate(first_admit_age = (as.numeric(floor_date(first_admittime, unit="day") - 
                                       floor_date(dob,    unit="day"), 
                            units="days")  /  365.242)  %>%  round(2))  %>%
  mutate(age_group = case_when(first_admit_age >  89 ~ ">89",
                               first_admit_age >= 14 ~ "adult",
                               first_admit_age <=  1 ~ "neonate",
                               TRUE                  ~ "middle"))
```

```{r, comment=NA}
nrow(dplyrResults)
```

```{r, comment=NA}
dplyrResults %>%
  head()     %>%
  Show()
```

# Example 5

Paul:  *Finally, we can use this information to categorize the patient’s that we
want in our cohort for further tests.*

## SQL

```{sql, connection=MimicDB, output.var="SQLresults"} 
WITH first_admission_time AS
(
  SELECT
      p.subject_id, p.dob, p.gender
      , MIN (a.admittime) AS first_admittime
      , MIN( ROUND( (cast(admittime as date) - cast(dob as date)) / 365.242,2) )
          AS first_admit_age
  FROM mimiciii.patients p
  INNER JOIN mimiciii.admissions a
  ON p.subject_id = a.subject_id
  GROUP BY p.subject_id, p.dob, p.gender
  ORDER BY p.subject_id
)
, age as
(
  SELECT
      subject_id, dob, gender,
      first_admittime, first_admit_age,
      CASE
          -- all ages > 89 in the database were replaced with 300
          -- we check using > 100 as a conservative threshold to ensure we capture all these patients
          WHEN first_admit_age > 100
               then '>89'
          WHEN first_admit_age >= 14
               THEN 'adult'
          WHEN first_admit_age <= 1
               THEN 'neonate'
          ELSE 'middle'
      END AS age_group
  FROM first_admission_time
)
select age_group, gender, 
       count(subject_id) as NumberOfPatients
from age
group by age_group, gender
```

```{r, comment=NA}
SQLresults %>%
  head()   %>%
  Show()
```

## dplyr

Extend previous results 

```{r, comment=NA}
dplyrResults                         %>%
  group_by(age_group, gender)        %>%
  summarize(NumberOfPatients = n())  %>%
  Show()
```

Use `spread` to show results by gender in more compact format

```{r, comment=NA}
byAgeGroupAndGender <-
  dplyrResults                       %>%
  group_by(age_group, gender)        %>%
  summarize(NumberOfPatients = n())  %>%
  ungroup()                          %>%
  spread(gender, NumberOfPatients) 

byAgeGroupAndGender %>% Show()
```

Add Row Totals

```{r}
byAgeGroupAndGender <- 
  byAgeGroupAndGender                           %>%
  mutate(Total = rowSums(.[-1], na.rm = TRUE))  
```

Add Column Totals

```{r, comment=NA}
byAgeGroupAndGender <- 
  bind_rows(byAgeGroupAndGender,
            bind_cols(age_group = "Total",
                      byAgeGroupAndGender %>%
                        summarize_if(is.numeric, sum, na.rm=TRUE)))
```


```{r}
byAgeGroupAndGender %>% Show(bigMark = ",")
```


**Close Database**

```{r, comment=NA}
dbDisconnect(MimicDB)
```

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
