---
title: "MIMIC-III First Look -- RPostgres package"
output:
  html_document:
    toc: yes
    number_sections: true
  html_notebook:
    toc: yes       
---

efg | 2018-08-12 | Updated 2018-12-23.

Summary:  RPostgres DOES NOT correctly show dates back to 1800 (only to 1969).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```                                 

# Setup

## Packages

```{r}
library(DBI)
library(RPostgres)
library(dbplyr, warn.conflicts=FALSE)    
library(tidyverse)
library(kableExtra)
```

## Helper function

Common formatting mostly for data.frames/tibbles below

```{r}
Show <- function(data)
{
  data            %>%
  kable("html")   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE, font_size=10)
}
```        

# Open database with RPostgres

```{r}
MimicDB <- dbConnect(RPostgres::Postgres(),
                     host     = "localhost",
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",
                     options  = "-c search_path=mimiciii")
dbListTables(MimicDB) 
```

# Admissions Table

```{r}
admissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
```

```{r}
admissions %>% 
  head(5)  %>%
  Show()
```

```{r}
admissions                                %>%
  group_by(admission_type)                %>%
  count()                                 %>%
  Show()
```

```{r}
admissions                                %>%
  group_by(insurance)                     %>%
  count()                                 %>%
  Show()
```

```{r}
admissions                     %>%
  group_by(admission_location) %>%
  count()                      %>%
  Show()
```

# Patients Table

```{r}
patients <- tbl(MimicDB, in_schema("mimiciii", "patients"))
```

```{r}
patients                  %>%
  arrange(dob)            %>%
  head(5)                 %>%
  collect()               %>%
  Show()
```

Note DOB fields above go back to 1800, but result INCORRECTLY shows 1969.

SQL chunk below DOES work with RPostgres package, but INCORRECTLY shows 1969 in dob fields.

```{sql, connection=MimicDB, output.var="SQLchunkOutput"}
SELECT *
FROM   patients
ORDER BY dob
LIMIT 5
```

```{r}
SQLchunkOutput %>% Show()
```

Oddly, the cast below reveals the 1969 years are really 1800.

```{sql, connection=MimicDB, output.var="SQLchunkOutput"}
SELECT subject_id, dob, cast(dob as date) AS dobDate
FROM   patients
ORDER BY dob
LIMIT 5
```

```{r}
SQLchunkOutput %>% Show()
```

# Close database

```{r, comment=NA}
dbDisconnect(MimicDB)
```

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:",
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                