---
title: "MIMIC-III Count Characters"
output: html_notebook
---

Create frequency counts of all characters in files to check whether unusual or unexpected characters are in any of the MIMIC-III files.

efg | 2018-07-07

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()  
```

"Character Count" programs in this directory:

* CharCount.exe - Windows program to count characters in file.  Drag and drop to process a file.  When possible show character with count.

* charcnt.exe - command line program for Linux or Windows.  Creates 16-by-16 matrix of character counts.

* charcnt2.exe - command line program for Linux of Windows.  Ouput vector of 256 character counts, along with total, in single line.

Note:  Source code for programs can be found on [GitHub](https://github.com/EarlGlynn/CharacterCount).

```{r}
library(readr)      # read_csv
library(kableExtra) # kable_styling
```

## Make list of .csv data files

```{r}
dataDIR <- "../../data/" 
files <- list.files(dataDIR, full.names=TRUE)
```

## Get character counts for all files using charcnt2

Append each row of results to countsFile

```{r}
countsFile <- "##SaveIt.txt"
if (file.exists(countsFile)) file.remove(countsFile)

for (i in 1:length(files))
{
  cat(i, files[i], "\n") 
  shell(paste0(getwd(), "/charcnt2 ", files[i], " >> ", countsFile))
}
```

## Cleanup results

```{r}
results <- readLines(countsFile)
results <- substr(results, nchar(dataDIR)+1, nchar(results))
results <- c(paste0("Decimal,", paste(0:255, collapse=","), ",Total"),
             results)
file.remove(countsFile)
```

## Save results

```{r}
filename <- "MIMIC-III-character-counts.csv" 
writeLines(results, filename)
```

Use Excel to "make pretty" manually in **MIMIC-III-character-counts.xlsx**.

Study column counts to see if unexpected patterns exist across multiple files.

## Show raw results 

Better viewed in Excel.

```{r}
read_csv(filename)                                                        %>%
  kable("html", format.args=list(big.mark=","))                           %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                full_width=FALSE, position="left")                        %>%
  scroll_box(height="500px", width="1200px")
```

Normally edit the .csv file above (saving as an .xlsx file) to clarify that "Decimal" is the row label and that filenames are in the first column.  Add a TOTAL row for column sums.

Normally the Excel file is edited to show the first three rows as the Decimal, Hex and Character versions of the possible printable characters in a file.

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                      